---
title: "Data Visualization Project - Out-of-school rate per education level in Luxembourg"
name: "Oliver JACK"
data: 2023-11-21
---

### Setup

```{r}
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("forcats")
install.packages("rworldmap")
install.packages("gganimate")
```

```{r}
renv::restore()
renv::snapshot()
```

```{r}
library(tidyverse)
library(ggplot2)
library(forcats)
library(rworldmap)
library(gganimate)
```

### Tidying data set

```{r}
education <- read_csv("data/country_by_continent.csv", show_col_types = FALSE) |>
  inner_join(read_csv("data/education_data.csv", show_col_types = FALSE), join_by("Country" == "Country Name"))

education <- education |>
  pivot_longer(cols = c(4:ncol(education)),
               names_to = "Year",
               values_to = "Values")|>
  pivot_wider(names_from = Series,
              values_from = Values) |>
  filter(if_any(starts_with("Out-of-school"), ~!is.na(.))) |>
  mutate(Year = as.integer(Year)) |>
  rename("country" = `Country`,
         "continent" = Continent,
         "year" = Year,
         "gdp per capita (in $)" = `GDP per capita (current US$)`,
         "oos lower secondary (m & f)" = `Out-of-school adolescents of lower secondary school age, both sexes (number)`,
         "oos lower secondary (f)" = `Out-of-school adolescents of lower secondary school age, female (number)`,
         "oos lower secondary (m)" = `Out-of-school adolescents of lower secondary school age, male (number)`,
         "oos primary (m & f)" = `Out-of-school children of primary school age, both sexes (number)`,
         "oos primary (f)" = `Out-of-school children of primary school age, female (number)`,
         "oos primary (m)" = `Out-of-school children of primary school age, male (number)`,
         "oos upper secondary (m & f)" = `Out-of-school youth of upper secondary school age, both sexes (number)`,
         "oos upper secondary (f)" = `Out-of-school youth of upper secondary school age, female (number)`,
         "oos upper secondary (m)" = `Out-of-school youth of upper secondary school age, male (number)`,
         "tot. lower secondary (m & f)" = `School age population, lower secondary education, both sexes (number)`,
         "tot. lower secondary (f)" = `School age population, lower secondary education, female (number)`,
         "tot. lower secondary (m)" = `School age population, lower secondary education, male (number)`,
         "tot. primary (m & f)" = `School age population, primary education, both sexes (number)`,
         "tot. primary (f)" = `School age population, primary education, female (number)`,
         "tot. primary (m)" = `School age population, primary education, male (number)`,
         "tot. upper secondary (m & f)" = `School age population, upper secondary education, both sexes (number)`,
         "tot. upper secondary (f)" = `School age population, upper secondary education, female (number)`,
         "tot. upper secondary (m)" = `School age population, upper secondary education, male (number)`,
         "avg. years of schooling (m & f)" = `UIS: Mean years of schooling (ISCED 1 or higher), population 25+ years, both sexes`,
         "avg. years of schooling (f)" = `UIS: Mean years of schooling (ISCED 1 or higher), population 25+ years, female`,
         "avg. years of schooling (m)" = `UIS: Mean years of schooling (ISCED 1 or higher), population 25+ years, male`,
         "unemployment % (m & f)" = `Unemployment, total (% of total labor force)`,
         "unemployment % (f)" = `Unemployment, female (% of female labor force)`,
         "unemployment % (m)" = `Unemployment, male (% of male labor force)`) |>
  mutate(`oos % primary (f)` = `oos primary (f)`/`tot. primary (f)`,
         `oos % primary (m)` = `oos primary (m)`/`tot. primary (m)`,
         `oos % primary (m & f)` = `oos primary (m & f)`/`tot. primary (m & f)`,
         `oos % lower secondary (f)` = `oos lower secondary (f)`/`tot. lower secondary (f)`,
         `oos % lower secondary (m)` = `oos lower secondary (m)`/`tot. lower secondary (m)`,
         `oos % lower secondary (m & f)` = `oos lower secondary (m & f)`/`tot. lower secondary (m & f)`,
         `oos % upper secondary (f)` = `oos upper secondary (f)`/`tot. upper secondary (f)`,
         `oos % upper secondary (m)` = `oos upper secondary (m)`/`tot. upper secondary (m)`,
         `oos % upper secondary (m & f)` = `oos upper secondary (m & f)`/`tot. upper secondary (m & f)`,
         .after = year)
```

### Generating data subsets

```{r}
primary <- education |>
  filter(!is.na(`oos primary (m & f)`) & !is.na(`tot. primary (m & f)`)) |>
  select(country, continent, year, contains("primary"))

lower_secondary <- education |>
  filter(!is.na(`oos lower secondary (m & f)`) & !is.na(`tot. lower secondary (m & f)`)) |>
  select(country, continent, year, contains("lower secondary"))

upper_secondary <- education |>
  filter(!is.na(`oos upper secondary (m & f)`) & !is.na(`tot. upper secondary (m & f)`)) |>
  select(country, continent, year, contains("upper secondary"))
```

### Global out-of-school rate per education level

```{r}
primary_world <- primary |>
  group_by(year) |>
  summarise("primary" = sum(`oos primary (m & f)`)/sum(`tot. primary (m & f)`))

lower_secondary_world <- lower_secondary |>
  group_by(year) |>
  summarise("lower secondary" = sum(`oos lower secondary (m & f)`)/sum(`tot. lower secondary (m & f)`))

upper_secondary_world <- upper_secondary |>
  group_by(year) |>
  summarise("upper secondary" = sum(`oos upper secondary (m & f)`)/sum(`tot. upper secondary (m & f)`))

as_tibble(list(primary_world, lower_secondary_world, upper_secondary_world) |>
  reduce(full_join, by = "year") |>
  pivot_longer(!year, names_to = "level", values_to = "values")) |>
  ggplot(aes(x = year, y = values, colour = level)) +
  geom_point() +
  facet_wrap(~ factor(level, levels = c("primary", "lower secondary", "upper secondary"))) +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  labs(title = "Global out-of-school rate per education level",
       subtitle = "Annual average from 1970-2020") +
  scale_x_continuous(breaks = seq(1970, 2020, by = 10))
```

### Out-of-school rate per continent & education level

```{r}
primary_continent <- primary |>
  group_by(continent, year) |>
  summarise("primary" = sum(`oos primary (m & f)`)/sum(`tot. primary (m & f)`))

lower_secondary_continent <- lower_secondary |>
  group_by(continent, year) |>
  summarise("lower secondary" = sum(`oos lower secondary (m & f)`)/sum(`tot. lower secondary (m & f)`))

upper_secondary_continent <- upper_secondary |>
  group_by(continent, year) |>
  summarise("upper secondary" = sum(`oos upper secondary (m & f)`)/sum(`tot. upper secondary (m & f)`))

as_tibble(list(primary_continent, lower_secondary_continent, upper_secondary_continent) |>
  reduce(full_join, by = c("continent", "year")) |>
  pivot_longer(!c(continent, year), names_to = "level", values_to = "values")) |>
  mutate(level = fct_relevel(level, "primary", "lower secondary", "upper secondary")) |>
  ggplot(aes(x = year, y = values, colour = level)) +
  geom_point() +
  facet_grid(level ~ continent) +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  labs(title = "Out-of-school rate per continent & education level",
       subtitle = "Annual average from 1970-2020") +
  scale_x_continuous(breaks = c(1980, 2010))
```

```{r}
as_tibble(list(primary_continent, lower_secondary_continent, upper_secondary_continent) |>
  reduce(full_join, by = c("continent", "year")) |>
  pivot_longer(!c(continent, year), names_to = "level", values_to = "values")) |>
  mutate(level = fct_relevel(level, "primary", "lower secondary", "upper secondary")) |>
  ggplot(aes(x = continent, y = values, fill = level)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.title.x = element_blank(),
      axis.title.y = element_blank()) +
  labs(title = "Out-of-school rate per continent & education level",
     subtitle = "Average from 1970-2020",
     fill = "Education level")
```

### Countries with highest primary school out-of-school rate

```{r}
country_continent <- primary |>
  group_by(country, continent) |>
  summarise("primary" = sum(`oos primary (m & f)`)/sum(`tot. primary (m & f)`)) |>
  arrange(desc(primary)) |>
  head(n = 30)

country_continent |>
ggplot(aes(x = primary, y = fct_rev(fct_infreq(country, w = primary)), fill = continent)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Countries with highest primary school out-of-school rate",
       subtitle = "Average from 1970-2020",
       fill = "Continent") +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
```

### World map of primary school out-of-school rates (avg. from 1970-2020)

```{r}
primary_country <- primary |>
  group_by(country) |>
  summarise("primary" = sum(`oos primary (m & f)`)/sum(`tot. primary (m & f)`))

country_data <- joinCountryData2Map(primary_country, joinCode = "NAME", nameJoinColumn = "country")

par(mar=c(0, 0, 0, 0))

mapCountryData(country_data, 
nameColumnToPlot="primary", 
mapTitle="World map of primary school out-of-school rates (avg. from 1970-2020)",
colourPalette = "heat", 
oceanCol="lightblue", 
missingCountryCol="grey65", 
catMethod = "pretty")
```

### Annual male/female out-of-school ratio in primary school

```{r}
primary |>
  filter(!is.na(`oos primary (m)`) & !is.na(`oos primary (f)`)) |>
  group_by(year) |>
  summarise(male = sum(`oos primary (m)`),
            female = sum(`oos primary (f)`)) |>
  pivot_longer(!year,
               names_to = "gender",
               values_to = "proportion") |>
  ggplot(aes(x = proportion, y = fct_rev(as.factor(year)), fill = gender)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Annual male/female out-of-school ratio in primary school",
       fill = "Gender") +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  scale_y_discrete(breaks = as.character(seq(1970, 2020, by = 10)))
```

### Out-of-school rate per education level in Luxembourg

```{r}
primary_lux <- primary |>
  filter(country == "Luxembourg") |>
  group_by(year) |>
  summarise("primary" = sum(`oos primary (m & f)`)/sum(`tot. primary (m & f)`))

lower_secondary_lux <- lower_secondary |>
  filter(country == "Luxembourg") |>
  group_by(year) |>
  summarise("lower secondary" = sum(`oos lower secondary (m & f)`)/sum(`tot. lower secondary (m & f)`))

upper_secondary_lux <- upper_secondary |>
  filter(country == "Luxembourg") |>
  group_by(year) |>
  summarise("upper secondary" = sum(`oos upper secondary (m & f)`)/sum(`tot. upper secondary (m & f)`))

as_tibble(list(primary_lux, lower_secondary_lux, upper_secondary_lux) |>
  reduce(full_join, by = "year") |>
  pivot_longer(!year, names_to = "level", values_to = "values")) |>
  ggplot(aes(x = year, y = values, colour = level)) +
  geom_point() +
  facet_wrap(~ factor(level, levels = c("primary", "lower secondary", "upper secondary"))) +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(hjust = 0.7),
        axis.title.y = element_blank()) +
  labs(title = "Out-of-school rate per education level in Luxembourg",
       subtitle = "Annual average from 1970-2020") +
  scale_x_continuous(breaks = seq(1970, 2020, by = 10),
                     guide = guide_axis(n.dodge = 2))
```
