---
title: "Data Visualization Project - Global out-of-school rate analysis"
name: "Oliver JACK"
date: 2023-11-21
editor_options: 
chunk_output_type: inline
---

### Setup

```{r}
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("forcats")
install.packages("rworldmap")
install.packages("gganimate")
install.packages("broom")
install.packages("ggbeeswarm")
install.packages("ggthemes")
install.packages("viridis")
install.packages("gifski")
```

```{r}
renv::restore()
renv::snapshot()
```

```{r}
library(tidyverse)
library(ggplot2)
library(forcats)
library(rworldmap)
library(gganimate)
library(broom)
library(ggbeeswarm)
library(ggthemes)
library(viridis)
library(gifski)
```

```{r}
scale_colour_continuous <- scale_colour_viridis_c
scale_colour_discrete <- scale_colour_viridis_d
scale_colour_binned <- scale_colour_viridis_b

scale_fill_continuous <- scale_fill_viridis_c
scale_fill_discrete <- scale_fill_viridis_d
scale_fill_binned <- scale_fill_viridis_b

theme_set(theme_bw())
```

### Tidying data frame

```{r}
unfiltered_education <- read_csv("data/country_by_continent.csv", show_col_types = FALSE) |>
  inner_join(read_csv("data/education_data.csv", show_col_types = FALSE), join_by("Country" == "Country Name")) |>
  pivot_longer(cols = matches("^\\d{4}$"),
               names_to = "year",
               values_to = "values") |>
  pivot_wider(names_from = Series,
              values_from = values) |>
  mutate(year = as.integer(year)) |>
  rename(country = `Country`,
         continent = Continent,
         gdp_pc = `GDP per capita (current US$)`,
         oos_lower = `Out-of-school adolescents of lower secondary school age, both sexes (number)`,
         oos_lower_f = `Out-of-school adolescents of lower secondary school age, female (number)`,
         oos_lower_m = `Out-of-school adolescents of lower secondary school age, male (number)`,
         oos_primary = `Out-of-school children of primary school age, both sexes (number)`,
         oos_primary_f = `Out-of-school children of primary school age, female (number)`,
         oos_primary_m = `Out-of-school children of primary school age, male (number)`,
         oos_upper = `Out-of-school youth of upper secondary school age, both sexes (number)`,
         oos_upper_f = `Out-of-school youth of upper secondary school age, female (number)`,
         oos_upper_m = `Out-of-school youth of upper secondary school age, male (number)`,
         tot_lower = `School age population, lower secondary education, both sexes (number)`,
         tot_lower_f = `School age population, lower secondary education, female (number)`,
         tot_lower_m = `School age population, lower secondary education, male (number)`,
         tot_primary = `School age population, primary education, both sexes (number)`,
         tot_primary_f = `School age population, primary education, female (number)`,
         tot_primary_m = `School age population, primary education, male (number)`,
         tot_upper = `School age population, upper secondary education, both sexes (number)`,
         tot_upper_f = `School age population, upper secondary education, female (number)`,
         tot_upper_m = `School age population, upper secondary education, male (number)`,
         avg_yos = `UIS: Mean years of schooling (ISCED 1 or higher), population 25+ years, both sexes`,
         avg_yos_f = `UIS: Mean years of schooling (ISCED 1 or higher), population 25+ years, female`,
         avg_yos_m = `UIS: Mean years of schooling (ISCED 1 or higher), population 25+ years, male`,
         unemployment_p = `Unemployment, total (% of total labor force)`,
         unemployment_p_f = `Unemployment, female (% of female labor force)`,
         unemployment_p_m = `Unemployment, male (% of male labor force)`)

education <- unfiltered_education |>
  filter(if_any(starts_with("oos"), ~!is.na(.)))
```

### Adding aggregated columns to data frame

```{r}
rate_calculator <- function(row){
  numerator <- 0
  denominator <- 0
  if (!is.na(row[["oos_primary"]]) & !is.na(row[["tot_primary"]])) {
    numerator <- numerator + row[["oos_primary"]]
    denominator <- denominator + row[["tot_primary"]]
  }
  if (!is.na(row[["oos_lower"]]) & !is.na(row[["tot_lower"]])) {
    numerator <- numerator + row[["oos_lower"]]
    denominator <- denominator + row[["tot_lower"]]
  }
  if (!is.na(row[["oos_upper"]]) & !is.na(row[["tot_upper"]])) {
    numerator <- numerator + row[["oos_upper"]]
    denominator <- denominator + row[["tot_upper"]]
  }
  return(c(if (numerator != 0 && denominator != 0) c(numerator, denominator) else c(NA, NA)))
}

education$oos <- NA
education$tot <- NA
education$oos_rate <- NA

for (i in 1:nrow(education)) {
  result <- rate_calculator(education[i, ])
  education$oos[i] <- result[1]
  education$tot[i] <- result[2]
  education$oos_rate[i] <- ifelse(!is.na(result[1]), result[1]/result[2], NA)
}

# I was unable to implement it using the `apply` function.
```

### Function generating data subsets

```{r}
summariser <- function(level, ..., gender = FALSE){
  group <- enquos(...)
  if (level == "primary") {
    if (!gender) {
      return(education |>
               filter(!is.na(oos_primary) & !is.na(tot_primary)) |>
               group_by(!!!group) |>
               summarise(primary = sum(oos_primary)/sum(tot_primary)))
    }
    return(education |>
             filter(!is.na(oos_primary_m) & !is.na(oos_primary_f) & !is.na(tot_primary_m) & !is.na(tot_primary_f)) |>
             group_by(!!!group) |>
             summarise(primary_male = sum(oos_primary_m)/sum(tot_primary_m), primary_female = sum(oos_primary_f)/sum(tot_primary_f)))
  }
  if (level == "lower") {
    if (!gender) {
      return(education |>
               filter(!is.na(oos_lower) & !is.na(tot_lower)) |>
               group_by(!!!group) |>
               summarise(lower = sum(oos_lower)/sum(tot_lower)))
    }
    return(education |>
             filter(!is.na(oos_lower_m) & !is.na(oos_lower_f) & !is.na(tot_lower_m) & !is.na(tot_lower_f)) |>
             group_by(!!!group) |>
             summarise(lower_male = sum(oos_lower_m)/sum(tot_lower_m), lower_female = sum(oos_lower_f)/sum(tot_lower_f)))
  }
  if (level == "upper") {
    if (!gender) {
      return(education |>
               filter(!is.na(oos_upper) & !is.na(tot_upper)) |>
               group_by(!!!group) |>
               summarise(upper = sum(oos_upper)/sum(tot_upper)))
    }
    return(education |>
             filter(!is.na(oos_upper_m) & !is.na(oos_upper_f) & !is.na(tot_upper_m) & !is.na(tot_upper_f)) |>
             group_by(!!!group) |>
             summarise(upper_male = sum(oos_upper_m)/sum(tot_upper_m), upper_female = sum(oos_upper_f)/sum(tot_upper_f)))
  }
}
```

### Out-of-school rate by continent & education level

```{r}
summariser("primary", continent, year) |>
  full_join(summariser("lower", continent, year), by = c("continent", "year")) |>
  full_join(summariser("upper", continent, year), by = c("continent", "year")) |>
  pivot_longer(!c(continent, year), names_to = "level", values_to = "values") |>
  ggplot(aes(x = year, y = values, colour = continent)) +
  geom_smooth() +
  facet_grid(factor(level,
                    levels = c("primary", "lower", "upper"),
                    labels = c("primary", "lower secondary", "upper secondary")) ~ continent) +
  labs(title = "Out-of-school rate by continent & education level",
       subtitle = "Annual average from 1970-2020") +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  scale_x_continuous(breaks = c(1980, 2010)) +
  scale_y_continuous(labels = scales::percent_format(scale = 100))
```

### Animation of global map with primary out-of-school rates (annual average from 1970-2020)

```{r}
data <- unfiltered_education |>
  mutate(primary = ifelse(is.na(oos_primary) | is.na(tot_primary), NA, oos_primary / tot_primary)) |>
  select(country, continent, year, primary)

getMap(resolution = "low") |>
  spTransform(CRS("+proj=robin")) |>
  fortify() |>
  left_join(data, by = c("id" = "country"), relationship = "many-to-many") |>
  ggplot() +
  geom_polygon(aes(x = long, y = lat, group = group, fill = primary), color = "grey30") +
  scale_fill_viridis(option = "F", begin = 1, end = 0, limits = c(0, 1),
                     breaks = c(0, 0.5, 1), na.value = "grey85",
                     labels = scales::percent_format(scale = 100)) +
  theme_void() +
  labs(title = "Primary out-of school rate by country in {frame_time}") +
  theme(plot.title = element_text(size = 14),
        legend.title = element_blank(),
        legend.position = "bottom") +
  transition_time(year) +
  ease_aes("linear") -> o

# animate(o, renderer = gifski_renderer())
# anim_save("worldmap.gif")
```

### Top 30 countries with the highest primary out-of-school rate from 2010-2019

```{r}
data1 <- education |>
  filter(year >= 2010 & year <= 2019 & !is.na(oos_primary))

world_avg <- sum(data1$oos_primary) / sum(data1$tot_primary)

data1 |>
  group_by(country, continent) |>
  summarise(primary = sum(oos_primary) / sum(tot_primary)) |>
  arrange(desc(primary)) |>
  head(n = 30) |>
  ggplot(aes(x = primary, y = fct_reorder(country, primary), fill = continent)) + 
  geom_bar(stat = "identity") + 
  geom_vline(xintercept = world_avg, linetype = "dashed", colour = "white") +
  geom_text(aes(x = world_avg + 0.05, label = "world avg.", y = 15.2), size = 3.5, colour = "white") +
  labs(title = "Top 30 countries with the highest primary out-of-school rate",
       subtitle = "Average from 2010-2019",
       fill = "Continent") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = c(0.85, 0.2)) +
  scale_x_continuous(labels = scales::percent_format(scale = 100),
                     breaks = c(world_avg, seq(0, 1, by = 0.20)))
```

### Distribution of out-of-school rate by continent & gender

```{r}
summariser("primary", country, continent, year, gender = TRUE) |>
  filter((year >= 1980 & year <= 1989) | (year >= 2010 & year <= 2019)) |>
  pivot_longer(cols = starts_with("primary"),
               names_to = "gender",
               values_to = "values",
               names_prefix = "primary_") |>
  mutate(year_group = ifelse(year >= 1980 & year <= 1989, "1980-1989", "2010-2019")) |>
  ggplot(aes(x = factor(year_group), y = values, colour = gender)) +
  geom_quasirandom(dodge.width = 0.7) +
  facet_wrap(~ factor(continent)) +
  labs(title = "Distribution of primary out-of-school rate by continent & gender",
       subtitle = "Comparison between 1980-1989 & 2010-2019",
       colour = "Gender") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(scale = 100))
```

### Annual male/female primary out-of-school ratio per continent

```{r}
summariser("primary", continent, year, gender = TRUE) |>
  mutate(female_proportion = primary_female / (primary_male + primary_female)) |>
  ggplot(aes(x = factor(year), y = female_proportion, colour = continent)) +
  geom_smooth(aes(group = continent), se = FALSE) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  facet_wrap(~ continent) +
  labs(title = "Female proportion of primary out-of school students per continent",
       subtitle = "Annual average from 1970-2020",
       colour = "Continent") +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  scale_x_discrete(breaks = seq(1975, 2015, by = 10)) +
  scale_y_continuous(labels = scales::percent_format(scale = 100))
```

### Top 10 countries with highest out-of-school rate gender disparity

```{r}
education |>
  filter(year >= 2010 & year <= 2019 & !is.na(oos_primary) & !is.na(oos_primary_f) & !is.na(oos_primary_m)) |>
  group_by(country, continent) |>
  summarise(primary = sum(oos_primary) / sum(tot_primary), primary_f = sum(oos_primary_f) / sum(tot_primary_f), primary_m = sum(oos_primary_m) / sum(tot_primary_m)) |>
  mutate(delta = abs(primary_f - primary_m)) |>
  arrange(desc(delta)) |>
  head(10) |>
  pivot_longer(starts_with("primary"), names_to = "gender", values_to = "values") |>
  ggplot(aes(x = values, y = fct_reorder(country, delta))) +
  geom_line(linewidth = 1, alpha = 0.5) +
  geom_point(aes(shape = factor(gender,
                      levels = c("primary_f", "primary_m", "primary"),
                      labels = c("female", "male", "overall")),
                 colour = delta), size = 4) +
  geom_vline(xintercept = world_avg, linetype = "dashed") +
  geom_text(aes(x = world_avg + 0.06, label = "world avg.", y = 4), size = 3.5) +
  labs(title = "Top 10 countries with highest out-of-school rate gender disparity",
       subtitle = "Average from 2010-2019",
       colour = "% difference",
       shape = "Gender") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.minor.x = element_blank()) +
  scale_x_continuous(labels = scales::percent_format(scale = 100),
                     breaks = c(world_avg, seq(0.2, 1, by = 0.1))) +
  scale_colour_continuous(labels = scales::percent_format(scale = 100))
```

### R<sup>2</sup> value for linear model of each country

```{r}
models <- education |>
  group_by(country) |>
  filter(n() >= 20) |>
  ungroup() |>
  nest_by(country, continent) |>
  mutate(model = list(lm(oos_rate ~ year, data = data)),
         coeff = coef(model)[2],
         rsq = pluck(glance(model), "r.squared"))

models |>
  ggplot(aes(x = rsq, y = fct_reorder(country, rsq))) +
  geom_point(aes(colour = continent), 
             alpha = 0.7, size = 2) +
  labs(title = expression(R^2 ~ "value for linear model of each country"),
     subtitle = "Only considering countries that have reported for over 20 years",
     x = expression(R^2),
     y = "Country",
     color = "Continent",
     parse = TRUE) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = c(0.87, 0.3)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.2))
```

### Out-of-school rate by country with highest R<sup>2</sup> value

```{r}
models |>
  filter(rsq > 0.8) |>
  arrange(desc(rsq)) |>
  head(10) |>
  ungroup() |>
  unnest(data) |>
  ggplot(aes(x = year, y = oos_rate, colour = country)) +
  geom_point() +
  geom_smooth(method = "lm", formula = 'y ~ x', se = FALSE, linewidth = 0.8) +
  facet_wrap(vars(country), nrow = 2) +
  labs(title = expression("Out-of-school rate by country with highest " ~ R^2 ~ "value"),
       subtitle = "Annual average from 1970-2020",
       parse = TRUE) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none") +
  scale_x_continuous(breaks = c(1980, 2010)) +
  scale_y_continuous(labels = scales::percent_format(scale = 100))
```

### Animation of correlation between out-of-school rate & GDP per capita from 1970-2018

```{r}
education |>
  filter(!is.na(oos_rate) & !is.na(gdp_pc)) |>
  ggplot(aes(x = gdp_pc, y = oos_rate, colour = continent, size = tot)) +
  geom_point() +
  scale_size(range = c(2, 12)) +
  labs(title = "Correlation between out-of-school rate & GDP per capita in {frame_time}", 
       x = "GDP per capita", 
       y = "Out-of school rate") +
  theme(plot.title = element_text(size = 14),
        legend.position = "none") +
  scale_x_log10(labels = scales::label_comma()) +
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  transition_time(year) +
  ease_aes("linear") -> p

animate(p, renderer = gifski_renderer())
anim_save("oos_gdp.gif")
```

### Animation of correlation between out-of-school rate & unemployment rate from 1970-2018

```{r}
education |>
  filter(!is.na(oos_rate) & !is.na(unemployment_p)) |>
  ggplot(aes(x = unemployment_p, y = oos_rate, colour = continent, size = tot)) +
  geom_point() +
  scale_size(range = c(2, 12)) +
  labs(title = "Correlation between out-of-school rate & unemployment rate in {frame_time}", 
       x = "Unemployment rate", 
       y = "Out-of school rate") +
  theme(plot.title = element_text(size = 14),
        legend.position = "none") +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  transition_time(year) +
  ease_aes("linear") -> q

animate(q, renderer = gifski_renderer())
anim_save("oos_unemployment.gif")
```
